---
title: "p8106_hw2_dd2948"
author: "David DeStephano"
date: "March 15, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r  message=FALSE, warning = FALSE}
library(caret)
library(splines)
library(lasso2) 
library(mgcv)
library(tidyverse)
library(ggplot2)
library(earth)
library(pdp)
```

#Import the data
```{r message=FALSE, warning = FALSE}
college<-read_csv("College.csv")
```

#Set seed to 2
```{r}
set.seed(2)
```


#Scatter plots of response vs. predictors.
```{r}
college %>%
  gather(-Outstate, -College, key = "var", value = "value") %>%
  ggplot(aes(x = value, y = Outstate)) +
    geom_point() +
    facet_wrap(~ var, scales = "free_x") +
    theme_bw()
```

#Alternative plot
```{r}
map(college, ~sum(is.na(.)))

college2<-college %>% select(-College) %>% select(-Outstate,Outstate) %>% filter(!is.na(Outstate))


# matrix of predictors
x <-model.matrix(Outstate~.,college2)[,-1]
# vector of response
y <- college2$Outstate


theme1 <-trellis.par.get()
theme1$plot.symbol$col <-rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <-rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <-rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(x, y, plot = "scatter", labels =c("","Y"),type =c("p"), layout =c(4, 2))
```


#b) Fit a smoothing spline model using Terminal as the only predictor of Outstate for a range of degrees of freedom, as well as the degree of freedom obtained by generalized cross-validation, and plot the resulting fits. Describe the results obtained.

```{r}
fit.ss <-smooth.spline(college2$Terminal, college2$Outstate)
fit.ss$df

Terminallims <-range(college2$Terminal)
Terminal.grid <-seq(from = Terminallims[1],to = Terminallims[2])

pred.ss <-predict(fit.ss,
                  x = Terminal.grid)

pred.ss.df <-data.frame(pred = pred.ss$y,
                        Terminal = Terminal.grid)

p <-ggplot(data = college2,aes(x = Terminal, y =Outstate)) +
  geom_point(color =rgb(.2, .4, .2, .5))

p+geom_line(aes(x = Terminal, y = pred), data = pred.ss.df,color =rgb(.8, .1, .1, 1))+ theme_bw()
```


#(c) Fit a generalized additive model (GAM) using all the predictors. Plot the results and explain your findings.
```{r}
gam.m1 <-gam(Outstate~Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+P.Undergrad+Room.Board+Books+ Personal+PhD+Terminal+S.F.Ratio+perc.alumni+Expend+Grad.Rate, data = college2)


gam.m2 <-gam(Outstate~Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+P.Undergrad+Room.Board+Books+ Personal+PhD+s(Terminal)+S.F.Ratio+perc.alumni+Expend+Grad.Rate, data = college2)


gam.m3 <-gam(Outstate~Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+P.Undergrad+te(Room.Board, Personal)+Books+ PhD+s(Terminal)+S.F.Ratio+perc.alumni+Expend+Grad.Rate, data = college2)


gam.m3 <-gam(Outstate~Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+P.Undergrad+te(Room.Board, Personal)+Books+ PhD+s(Terminal)+S.F.Ratio+perc.alumni+Expend+Grad.Rate, data = college2)


gam.m4 <-gam(Outstate~Apps+te(Accept, Enroll)+Top10perc+Top25perc+F.Undergrad+P.Undergrad+te(Room.Board, Personal)+Books+ PhD+s(Terminal)+S.F.Ratio+perc.alumni+Expend+Grad.Rate, data = college2)


anova(gam.m1, gam.m2, gam.m3, gam.m4, test = "F")


plot(gam.m2)

vis.gam(gam.m3, view =c("Room.Board","Personal"),plot.type = "contour", color = "topo")

plot(gam.m4)
vis.gam(gam.m4, view =c("Accept","Enroll"),plot.type = "contour", color = "topo")


```

According to the anova procedure, model 4 is the best fitting model



#(d) Fit a multivariate adaptive regression spline (MARS) model using all the predictors. Report the final model. Present the partial dependence plot of an arbitrary predictor in your final model.

```{r}
ctrl1 <-trainControl(method = "cv", number = 10)
mars_grid <-expand.grid(degree = 1:4,
                        nprune =2:25)

mars.fit <-train(x, y,
                 method = "earth",
                 tuneGrid = mars_grid,
                 trControl = ctrl1)

ggplot(mars.fit)

mars.fit$bestTune

coef(mars.fit$finalModel)
```

Partial Dependence plot
```{r}
p1 <-partial(mars.fit, pred.var =c("Room.Board"), grid.resolution = 10)%>% autoplot()

p2 <-partial(mars.fit, pred.var =c("Room.Board","Personal"), grid.resolution = 10)%>%plotPartial(levelplot = FALSE, zlab = "yhat", drape = TRUE,screen =list(z = 20, x = -60))

grid.arrange(p1, p2, ncol = 2)
```


#Compare to caret GAM

```{r}
gam.fit <-train(x, y,
                method = "gam",
                tuneGrid =data.frame(method = "GCV.Cp", select =c(TRUE,FALSE)),
                trControl = ctrl1)


bwplot(resamples(list(mars = mars.fit,gam = gam.fit)), metric = "RMSE")

gam.m4

```



```{r}
resamp <- resamples(list(gam=gam.fit,
                         ridge=mars.fit))

summary(resamp)
```

#(e) Based on the above GAM and MARS models, predict the out-of-state tuition of Columbia University.


```{r}
columbia<-college %>% filter(College=="Columbia University")
pred.gam <-predict(gam.fit, newdata =columbia)
pred.mars <-predict(mars.fit, newdata =columbia)

#predicted using the MGCV GAM instead of Caret
pred.mgcvgam=predict(gam.m4,newdata=columbia,se = TRUE)


pred.gam
pred.mars
pred.mgcvgam
```
The predicted value for the GAM model is $17728.51, and the predicted value for the MARs model is $18456.89
The predicted value using the fourth MGCV GAM model was 19240.62







For b) not sure if plotted both with range of DF and CV DF
